{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Models for detecting BHBH, BHNS and NSNS binaries with LISA

### 2021-11-25 paper/tex/sections/introduction.tex
introduction proofread

The detection of NSNSs in LISA could improve our understanding of many phenomena. Galactic NSNSs have been observed with electromagnetic signals for several decades (e.g.\ \citealp{Hulse+1975, Antoniadis+2016}, see also refs.\ in \citealp{Tauris+2017}) and more recently the mergers of NSNS binaries have been detected with ground-based gravitational-wave detectors \citep[e.g.][]{Abbott+2017_NSNS}. A LISA detectable NSNS with a pulsar component close to merger would be ideal for connecting these populations, as the binary could be observed from inspiral to merger. NSNS (and possible BHNS) binaries are useful sources for understanding the origin of r-process elements \citep[e.g.][]{Eichler+1989} as well as the electromagnetic counterparts to gravitational-wave signals, such as kilonovae \citep[e.g.][]{Li+1998, Metzger+2017}, short gamma-ray bursts \citep[e.g.][]{Berger+2014}, radio emission \citep[e.g.][]{Hotokezaka+2016} and neutrinos \citep[e.g.][]{Kyutoku+2018}.

@@

One particularly interesting class of potential LISA sources is BHNSs. With the recent detection of two BHNSs by the LIGO scientific collaboration, the existence of these DCOs has been confirmed \citep{TheLIGOScientificCollaboration+2021}. However, with only two detections (not including the low-confidence candidate GW190426 \citealt{Abbott+2020_GWTC2} or GW190425, GW190814 and GW190917 which have not been ruled out as BHNSs \citealt{Abbott+2020_GW190425,Abbott+2020_GW190814, GWTC_2_1} ) and no electromagnetic counterparts, the formation rate and properties of BHNSs are still uncertain. Current predictions for the merger rate of BHNSs range across three orders of magnitude \citep[e.g.][]{Abadie+2010, Mandel+2021} so the number of detections in LISA will be important in reducing this uncertainty, thereby refining our understanding of the remnants and evolution of massive stars. Similar to NSNSs, these binaries are also expected to have electromagnetic counterparts. A distinctly exciting possibility is the detection of a pulsar--BH system or millisecond pulsar--BH system \citep{Narayan+1991, Pol+2021}. These systems could be observed not only by LISA, but also radio telescopes such as MeerKAT and the Square Kilometre Array (SKA, \citealt{Dewdney+2009}), which would help to improve the measurement of individual system parameters and to constrain uncertain binary evolution processes \citep[e.g.][]{Pfahl+2005,Chattopadhyay+2020}.

@@

In this paper, we present models for the detection rate and distribution of binary properties (masses, frequency, eccentricity, distance, merger time) of BHBH, BHNS and NSNS binaries formed through isolated binary evolution in the Milky Way. We explore the effect of varying physical assumptions in our population synthesis model on our results as well as discuss the effect of extending the LISA mission length and the prospects for distinguishing DCO detections from the WDWD background.

Earlier work on BHBHs, BHNSs and NSNSs in LISA has used a variety of population synthesis codes, Milky Way models and LISA specifications, resulting in a wide range of predictions \citep{Nelemans+2001,Belczynski+2010,Liu+2014,Lamberts+2019,Lau+2020,Breivik+2020,Sesana+2020, Shao+2021}. We build upon previous efforts but with several important improvements. We explore the effects of varying binary physics assumptions by repeating our analysis for \nModels{} different models and comparing the effects on the detection rate and distributions of source parameters \citefloorp{}. We use a model for the Milky Way that accounts for the chemical enrichment history and is calibrated on the APOGEE survey \citep{Majewski+2017,Frankel+2018}, whereas most others did not consider the effect of metallicity in detail (see however \citealp{Lamberts+2019, Sesana+2020}). We provide a detailed treatment of the eccentricity of detectable sources, both for the inspiral evolution as well as gravitational wave signal during the LISA mission. Moreover, our binary population synthesis simulation is the most extensive of its kind to date and makes use of the adaptive sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019, Broekgaarden+2021}. Overall we simulate over 2 billion massive binaries to produce the DCO populations used in this work. We find that this large number of simulations is important to reduce the sampling noise even when using adaptive importance sampling.

@@

Our paper is structured as follows. In Section~\ref{sec:method}, we describe our methods for synthesising a population of binaries, the variations of physical assumptions that we consider, how we simulate the Milky Way distribution of DCOs and our methods for calculating a detection rate for LISA. We present the main results for our fiducial model in Section~\ref{sec:results}, before exploring the variations in the detectable population when changing our physical assumptions in Section~\ref{sec:variations}. In Section~\ref{sec:discussion} we discuss these results. In Section~\ref{sec:compare_studies}, we compare and contrast our methods and findings to previous work and finish with our conclusions in Section~\ref{sec:conclusion}.

>>>>

The detection of NSNSs in LISA could improve our understanding of many phenomena. Galactic NSNSs have been observed with electromagnetic signals for several decades (e.g.\ \citealp{Hulse+1975, Antoniadis+2016}, see also refs.\ in \citealp{Tauris+2017}) and more recently the mergers of NSNSs have been detected with ground-based gravitational-wave detectors \citep[e.g.][]{Abbott+2017_NSNS}. A LISA detectable NSNS with a pulsar component close to merger would be ideal for connecting these populations, as the binary could be observed from inspiral to merger. NSNSs (and possibly BHNSs) are useful sources for understanding the origin of r-process elements \citep[e.g.][]{Eichler+1989} as well as the electromagnetic counterparts to gravitational-wave signals, such as kilonovae \citep[e.g.][]{Li+1998, Metzger+2017}, short gamma-ray bursts \citep[e.g.][]{Berger+2014}, radio emission \citep[e.g.][]{Hotokezaka+2016} and neutrinos \citep[e.g.][]{Kyutoku+2018}.

@@

One particularly interesting class of potential LISA sources is BHNSs. With the recent detection of two BHNSs by the LIGO scientific collaboration, the existence of these DCOs has been confirmed \citep{TheLIGOScientificCollaboration+2021}. However, with only two detections (not including the low-confidence candidate GW190426, \citealt{Abbott+2020_GWTC2}, or GW190425, GW190814 and GW190917 which have not been ruled out as BHNSs, \citealt{Abbott+2020_GW190425,Abbott+2020_GW190814, GWTC_2_1}) and no electromagnetic counterparts, the formation rate and properties of BHNSs are still uncertain. Current predictions for the merger rate of BHNSs range across three orders of magnitude \citep[e.g.][]{Abadie+2010, Mandel+2021} so the number of detections in LISA will be important in reducing this uncertainty, thereby refining our understanding of the remnants and evolution of massive stars. Similar to NSNSs, these binaries are also expected to have electromagnetic counterparts. A distinctly exciting possibility is the detection of a pulsar--BH system or millisecond pulsar--BH system \citep{Narayan+1991, Pol+2021}. These systems could be observed not only by LISA, but also radio telescopes such as MeerKAT and the Square Kilometre Array (SKA, \citealt{Dewdney+2009}), which would help to improve the measurement of individual system parameters and to constrain uncertain binary evolution processes \citep[e.g.][]{Pfahl+2005,Chattopadhyay+2020}.

@@

In this paper, we present models for the detection rate and distribution of binary properties (masses, frequency, eccentricity, distance, merger time) of BHBHs, BHNSs and NSNSs formed through isolated binary evolution in the Milky Way. We explore the effect of varying physical assumptions in our population synthesis model on our results as well as discuss the effect of extending the LISA mission length and the prospects for distinguishing DCO detections from the WDWD background.

Earlier work on BHBHs, BHNSs and NSNSs in LISA has used a variety of population synthesis codes, Milky Way models and LISA specifications, resulting in a wide range of predictions \citep{Nelemans+2001,Belczynski+2010,Liu+2014,Lamberts+2019,Lau+2020,Breivik+2020,Sesana+2020, Shao+2021}. We build upon previous efforts but with several important improvements. We explore the effects of varying binary physics assumptions by repeating our analysis for \nModels{} different models and comparing the effects on the detection rate and distributions of source parameters \citefloorp{}. We use a model for the Milky Way that accounts for the chemical enrichment history and is calibrated on the APOGEE survey \citep{Majewski+2017,Frankel+2018}, whereas most others did not consider the effect of metallicity in detail (see however \citealp{Lamberts+2019, Sesana+2020}). We provide a detailed treatment of the eccentricity of detectable sources, both for the inspiral evolution as well as gravitational wave signal during the LISA mission. Moreover, the grid of binary population synthesis simulations that we use is the most extensive of its kind to date and makes use of the adaptive sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019, Broekgaarden+2021}. Overall over 2 billion massive binaries were simulated to produce the DCO populations used in this work. We find that this large number of simulations is important to reduce the sampling noise even when using adaptive importance sampling.

@@

Our paper is structured as follows. In Section~\ref{sec:method}, we describe the methods for synthesising a population of binaries, the variations of physical assumptions that we consider, how we simulate the Milky Way distribution of DCOs and our methods for calculating a detection rate for LISA. We present the main results for our fiducial model in Section~\ref{sec:results}, before exploring the variations in the detectable population when changing our physical assumptions in Section~\ref{sec:variations}. In Section~\ref{sec:discussion} we discuss these results. In Section~\ref{sec:compare_studies}, we compare and contrast our methods and findings to previous work and finish with our conclusions in Section~\ref{sec:conclusion}.
